Generalized Measurement Invariance Tests for Factor Analysis

### 2012-01-26 www/psychoco2012/invariance.tex
Made small changes to lavaan-S3.R so that it works with the most recent devel version of lavaan (0.4-12): two slots in lavaan objects were moved.
Revised Psychoco 2012 slides:
  - Divided stereotype threat results into multiple graphs
  - Other minor changes to equations


git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@90 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% FA path diagram, types of MI
\frame{
  \frametitle{Studying Measurement Invariance: Age <= 16}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{model_under16.pdf}
  \end{center}}

\frame{
  \frametitle{Studying Measurement Invariance: Age > 16}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{model_over16.pdf}
  \end{center}}

@@

\frame{
  \frametitle{Lack of Grouping}
  \begin{itemize}
    \item If we did not know the groups in advance, we could conduct a LRT wrt to all possible groups, taking the maximum.  The proposed tests are related to this idea.
  \end{itemize}
 % TODO Insert Fig 3
}

@@

\frame{
  \frametitle{Aggregating Scores}
 % define cumulative scores and other notation
  \begin{itemize}
    \item We need a way to aggregate scores across people so that we
      can draw some general conclusions.
      \begin{itemize}
        \item Order individuals by an auxiliary variable.\\ \ \\
        \item Define $t \in (1/n, n)$.
          The {\em empirical cumulative score process} is defined by:\\
      \end{itemize}
  \end{itemize}
  \begin{center}
    $\bm{B}(\hat{\bm{\theta}}; t) = \frac{1}{\sqrt{n}}
    \displaystyle\sum_{i=1}^{\lfloor nt 
      \rfloor} \psi ({\bm{x}}_i, \bm{\theta}).$\\ \ \\
  where $\lfloor nt \rfloor$ is the integer part of $nt$.
\end{center}}

@@

\frame{
  \frametitle{Theorem}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Theorem (Zeileis \& Hornik, 2007): Under the hypothesis of
      measurement invariance, the following functional central limit
      theorem holds:
  \begin{equation}
  ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) 
\overset{d}{\rightarrow} {\bm B}^{0}(\cdot),$
  \end{equation}
 where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
 information matrix and ${\bm B}^{0}(\cdot)$ is a
 $p$-dimensional Brownian bridge ($p$ is the number of parameters).
  \end{itemize}
}

@@

% TODO: Break up results over multiple slides
\frame{
  \frametitle{Results}
  \begin{center}
      \includegraphics[height=3in]{paper-gefp-4-1.pdf}
  \end{center}}

@@

\frame{
  \frametitle{Software}
  \begin{itemize}
    \item To carry out the tests, we utilize
      \begin{itemize}
        \item \texttt{lavaan} for model estimation.
        \item \texttt{estfun()} for score extraction, which is (currently) a combination of our own code and \texttt{lavaan} code.
        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
          \begin{itemize}
            \item Required input: Fitted model, function for score extraction, and information matrix (optional)
            \item \texttt{gefp()} constructs the process.
            \item \texttt{sctest()} and \texttt{plot()} calculate and visualize test statistics.
          \end{itemize}
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Current Work}
  \begin{itemize}
    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
    \item Detailed examination of test properties via simulation.\\ \ \\
    \item Extension to related psychometric issues (modification
      indices, mediation).
    \item Working paper: \url{http://econpapers.repec.org/RePEc:inn:wpaper:2011-09}
  \end{itemize}}

>>>>

% FA path diagram, types of MI
\frame{
  \frametitle{Studying Measurement Invariance: Age $\leq$ 16}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{model_under16.pdf}
  \end{center}}

\frame{
  \frametitle{Studying Measurement Invariance: Age $>$ 16}
  % path diagram illustrating parameters being the same for everyone
  \begin{center}
      \includegraphics[height=3in]{model_over16.pdf}
  \end{center}}

@@

\frame{
  \frametitle{Lack of Grouping}
  \begin{itemize}
    \item If we did not know the groups in advance, we could conduct a LRT wrt for each possible grouping, then take the maximum.  The proposed tests are related to this idea.
  \begin{center}
      \includegraphics[height=2.65in]{paper-lm-lr-19.pdf}
  \end{center}
  \end{itemize}}

@@

\begin{frame}[fragile]
  \frametitle{Aggregating Scores}
 % define cumulative scores and other notation
  \begin{itemize}
    \item We need a way to aggregate scores across people so that we
      can draw some general conclusions.
      \begin{itemize}
        \item Order individuals by an auxiliary variable.\\ \ \\
        \item Define $t \in (1/n, n)$.
          The {\em empirical cumulative score process} is defined by:\\
  \begin{equation*}
    \bm{B}(\hat{\bm{\theta}}; t) = \frac{1}{\sqrt{n}}
    \displaystyle\sum_{i=1}^{\lfloor nt 
      \rfloor} \psi ({\bm{x}}_i, \bm{\theta}).
    \end{equation*}
  where $\lfloor nt \rfloor$ is the integer part of $nt$.
      \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Theorem}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Theorem (Zeileis \& Hornik, 2007): Under the hypothesis of
      measurement invariance, the following functional central limit
      theorem holds:
  \begin{equation*}
  {\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) 
\overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
  \end{equation*}
 where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
 information matrix and ${\bm B}^{0}(\cdot)$ is a
 $p$-dimensional Brownian bridge ($p$ is the number of parameters).
  \end{itemize}
\end{frame}

@@

\frame{
  \frametitle{Results for Single Parameters}
  \begin{center}
      \includegraphics[height=3in]{gefp_4_ind.pdf}
  \end{center}}

\frame{
  \frametitle{Aggregated Results}
  \begin{center}
      \includegraphics[height=3in]{gefp_4_agg.pdf}
  \end{center}}

@@

\frame{
  \frametitle{Software}
  \begin{itemize}
    \item To carry out the tests, we utilize
      \begin{itemize}
        \item \texttt{lavaan} for model estimation.
        \item \texttt{estfun()} for score extraction, which is currently a combination of our own code and \texttt{lavaan} code.
        \item \texttt{strucchange} for carrying out the proposed tests with the scores.
          \begin{itemize}
            \item Required input: Fitted model, function for score extraction, and information matrix (optional)
            \item \texttt{gefp()} constructs the process.
            \item \texttt{sctest()} and \texttt{plot()} calculate and visualize test statistics.
          \end{itemize}
      \end{itemize}
  \end{itemize}}

\frame{
  \frametitle{Current Work}
  \begin{itemize}
    \item Continued test implementation via \texttt{strucchange} and \texttt{lavaan} (and possibly \texttt{OpenMx}).\\ \ \\
    \item Detailed examination of test properties via simulation.\\ \ \\
    \item Extension to related psychometric issues (modification
      indices, mediation).\\ \ \\
    \item Working paper: \url{http://econpapers.repec.org/RePEc:inn:wpaper:2011-09}
  \end{itemize}}
